---
title: "examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Examples of dynamet application

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r packages, include = FALSE}
library(terra)
library(dynamet)
library(dplyr)
library(ggplot2)
```


## 1. Minimal Example: Single Community Assembly

To understand the core lottery dynamics, we start with a single local community ($C=1$)
receiving individuals from a regional pool of 5 species.

```{r single-community}
# 1. Setup: 10 species in the regional pool
S <- 5
Meta.pool <- c(0.1, 0.2, 0.3, 0.3, 0.1) # Relative abundances

# 2. Parameters for a single community
# Js: Carrying capacity of 500 individuals
# M.migra: 1x1 matrix (closed system relative to other communities)
Js <- 500
M.migra <- matrix(1, nrow = 1, ncol = 1)
m.pool <- 0.05  # 5% immigration rate from the meta-pool
d.spp <- rep(1, S) # Uniform dispersal ability across species
nEpochs = 10 # Set the number of epochs

# 3. Initialize community with zero individuals
init_comm <- matrix(0, nrow = S, ncol = 1)

# 4. Run for 10 epochs
results <- dynamicMetacomm(
  nEpochs = nEpochs,
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = m.pool,
  d.spp = d.spp,
  lottery = TRUE,
  nIterations = 200,
  prop.dead.by.it = 0.1,
  Ea = 0.5,
  init.comm = init_comm,
  verbose = FALSE
)

# 5. Visualize the transition from zero to steady-state
sad_data <- do.call(rbind, lapply(1:nEpochs, function(i) {
  data.frame(
    Epoch = i,
    Species = factor(1:S),
    Abundance = as.vector(results[[i]])
  )
}))

# Plot: Stacked bar chart showing relative composition over time
ggplot(sad_data, aes(x = factor(Epoch), y = Abundance, fill = Species)) +
  geom_bar(stat = "identity", position = "fill") + # "fill" shows relative proportion
  labs(
    title = "Species Abundance Distribution (SAD) Over Time",
    x = "Epoch",
    y = "Relative Abundance"
  ) +
  theme_minimal()
```

## 2. Spatial Metacommunity Simulation

In this example, we model 100 communities arranged on a 10x10 grid, where carrying 
capacity (`Js`) varies spatially and migration is constrained by distance.

```{r spatial-sim, fig.width=8, fig.height=4}

# 1. Define the 10x10 Landscape
r <- rast(ncol=10, nrow=10, xmin=0, xmax=10, ymin=0, ymax=10)

# 2. Create spatially structured Carrying Capacity (Js)
values(r) <- 50 + 50 * sin(seq(0, pi, length.out=100))
Js <- as.vector(values(r))

# 3. Create Migration Matrix (M.migra)
coords <- crds(r)
dists <- as.matrix(dist(coords))
M.migra <- exp(-0.7 * dists)
M.migra <- M.migra / rowSums(M.migra)
diag(M.migra) <- 0 

# 4. Parameters
S <- 50
Meta.pool <- runif(S)
d.spp <- runif(S)

# 5. Run the Simulation
# Using masterEqMetacomm for a single epoch/iteration
community_abundance <- masterEqMetacomm(
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = 0.05,
  d.spp = d.spp,
  nIterations = 100,
  verbose = FALSE
)

# 6. Map results back to raster for visualization
alpha_div <- colSums(community_abundance > 0)
r_out <- r
values(r_out) <- alpha_div

# Visualize
par(mfrow=c(1,2))
plot(r, main = "Carrying Capacity (Js)", col = terrain.colors(10))
plot(r_out, main = "Alpha Diversity", col = cm.colors(10))
```

## 2. Spatial Metacommunity Simulation with a spatial filter

In this example, we model 100 communitieswith 20 species arranged on a 10x10 grid, where carrying 
capacity (`Js`) varies spatially and migration is constrained by distance.

```{r spatial-sim-filter, fig.width=8, fig.height=4}

# 1. Define the 10x10 Landscape Environment (e.g., elevation or pH)
env_r <- r <- rast(ncol=10, nrow=10, xmin=0, xmax=10, ymin=0, ymax=10)
values(env_r) <- seq(0, 1, length.out=100) # Simple gradient across the landscape
values(r) <- 50 + 50 * sin(seq(0, pi, length.out=100))
Js <- as.vector(values(r))

# 2. Define Species Niche Optima (S = 5)
S <- 20
niche_optima <- seq(0.5, 1, length.out = S)

# 3. Create the Environmental Filtering Matrix (FF)
# FF must be S x C (5 rows x 100 columns)
# We calculate a Gaussian penalty for each species in each cell
FF <- matrix(NA, nrow = S, ncol = 100)
env_vals <- values(env_r)

for(i in 1:S) {
  # Gaussian filter: 1 = perfect match, approaching 0 = high penalty
  FF[i, ] <- exp(-((env_vals - niche_optima[i])^2) / (2 * 0.2^2))
}

# 4. Run the simulation with Filtering
# Note: FF is now passed as a matrix of dimensions S x C
results <- masterEqMetacomm(
  Meta.pool = rep(1/S, S),
  Js = Js,
  M.migra = M.migra, # Assuming M.migra is defined as in the previous step
  m.pool = 0.05,
  d.spp = rep(1, S),
  FF = FF,           # Spatial Filter Applied
  nIterations = 500,
  verbose = FALSE
)

# 5. Visualize a species' distribution filtered by the environment
alpha_div <- colSums(ifelse(results > 0, 1, 0))
r_out <- r
values(r_out) <- alpha_div

# Visualize
par(mfrow=c(1,2))
plot(env_r, main = "Environment", col = terrain.colors(10))
plot(r_out, main = "Alpha Diversity", col = cm.colors(10))
```

## MISSING: example on how to include a custom updater function to dynamically change parameters across epochs














