---
title: "examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Examples of dynamet application

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r packages, message=FALSE }
library(terra)
library(dynamet)
library(dplyr)
library(ggplot2)
```


## 1. Single Community Assembly

To understand the core lottery dynamics, we start with a single local community ($C=1$)
receiving individuals from a regional pool of 5 species.

```{r single-community}
# 1. Setup: 10 species in the regional pool
S <- 5
Meta.pool <- c(0.1, 0.1, 0.2, 0.3, 0.5) # Relative abundances

# 2. Parameters for a single community
# Js: Carrying capacity of 500 individuals
# M.migra: 1x1 matrix (closed system relative to other communities)
Js <- 500
M.migra <- matrix(1, nrow = 1, ncol = 1) # Only one site considered
m.pool <- 0.05  # 1% immigration rate from the meta-pool
d.spp <- rep(0.5, S) # Uniform dispersal ability across species
nEpochs = 20 # Set the number of epochs

# 3. Initialize community with zero individuals
init_comm <- matrix( c(0.9, 0.9, 0.2, 0.2, 0.1), nrow = S, ncol = 1)

# 4. Run for 10 epochs
results <- dynamicMetacomm(
  nEpochs = nEpochs,
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = m.pool,
  d.spp = d.spp,
  lottery = TRUE,
  nIterations = 50,
  prop.dead.by.it = 0.05,
  Ea = 0.5,
  init.comm = init_comm,
  verbose = FALSE
)

# 5. Visualize the transition from zero to steady-state
sad_data <- do.call(rbind, lapply(1:nEpochs, function(i) {
  data.frame(
    Epoch = i,
    Epoch_label = paste0("Epoch_", i),
    Species = factor(1:S),
    Abundance = as.vector(results[[i]])
  )
}))

# 6. add the relative abundance for the starting point and from the species pool
sad_data <- rbind(
  data.frame(
    Epoch = 0,
    Epoch_label = "Start",
    Species = factor(1:S),
    Abundance = as.vector(init_comm)
  ),
  sad_data,
  data.frame(
    Epoch = nEpochs + 1,
    Epoch_label = "Pool",
    Species = factor(1:S),
    Abundance = Meta.pool 
  )
)


# Plot: Stacked bar chart showing relative composition over time
ggplot(sad_data, aes(x = factor(Epoch), y = Abundance, fill = Species)) +
  geom_bar(stat = "identity", position = "fill") + # "fill" shows relative proportion
  labs(
    title = "Species Abundance Distribution (SAD) Over Time",
    x = "Epoch",
    y = "Relative Abundance"
  ) +
  theme_minimal()
```

## 2. Spatial Metacommunity Simulation

In this example, we model 100 communities arranged on a 10x10 grid, where carrying 
capacity (`Js`) varies spatially and migration is constrained by distance.

```{r spatial-sim, fig.width=8, fig.height=4}

# 1. Define the 10x10 Landscape
r <- rast(ncol=10, nrow=10, xmin=0, xmax=10, ymin=0, ymax=10)

# 2. Create spatially structured Carrying Capacity (Js)
# Sinusoidal gradient across the landscape
Js_rast <- 20 + 50 * sin( init(r, "y") / 10 * pi)
Js <- as.vector(floor(values(Js_rast)))

# 3. Create Migration Matrix (M.migra)
coords <- crds(r)
dists <- as.matrix(dist(coords))
M.migra <- 10^(-0.01 * dists)

# 4. Parameters
S <- 100
Meta.pool <- runif(S)
d.spp <- runif(S)

# 5. Run the Simulation
# Using masterEqMetacomm for a single epoch/iteration
community_abundance <- masterEqMetacomm(
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = 0.05,
  d.spp = d.spp,
  nIterations = 500,
  verbose = FALSE
)

# 6. Map results back to raster for visualization
alpha_div <- colSums(community_abundance > 0)
r_out <- r
values(r_out) <- alpha_div

# Visualize
par(mfrow=c(2,2))
plot(Js_rast, main = "Carrying Capacity (Js)", col = terrain.colors(10))
plot(r_out, main = "Alpha Diversity", col = cm.colors(10))
plot(Js, alpha_div, xlab = "Carrying capacity", ylab = "Alpha diversity")

# Check with a linear model the relationship between alpha and Js
summary(lm(alpha_div ~ Js))

```

## 3. Spatial Metacommunity Simulation with a spatial filter

In this example, we model 100 communities with 20 species arranged on a 10x10 grid, where carrying 
capacity (`Js`) varies spatially and migration is constrained by distance.

```{r spatial-sim-filter, fig.width=8, fig.height=4}

# 1. Define the 10x10 Landscape Environment (e.g., elevation or pH)
env_r <- r <- rast(ncol=10, nrow=10, xmin=0, xmax=10, ymin=0, ymax=10)
values(env_r) <- seq(0, 1, length.out=100) # Simple gradient across the landscape
values(r) <- 50 + 50 * sin(seq(0, pi, length.out=100))
Js <- as.vector(floor(values(r)))

# 2. Define Species Niche Optima (S = 5)
S <- 50
niche_optima <- seq(0.75, 1, length.out = S)

# 3. Create the Environmental Filtering Matrix (FF)
# FF must be S x C (5 rows x 100 columns)
# We calculate a Gaussian penalty for each species in each cell
FF <- matrix(NA, nrow = S, ncol = 100)
env_vals <- values(env_r)

for(i in 1:S) {
  # Gaussian filter: 1 = perfect match, approaching 0 = high penalty
  FF[i, ] <- exp(-((env_vals - niche_optima[i])^2) / (2 * 0.2^2))
}

# 4. Run the simulation with Filtering
# Note: FF is now passed as a matrix of dimensions S x C
results <- masterEqMetacomm(
  Meta.pool = rep(1/S, S),
  Js = Js,
  M.migra = M.migra, # Assuming M.migra is defined as in the previous step
  m.pool = 0.05,
  d.spp = rep(1, S),
  FF = FF,           # Spatial Filter Applied
  nIterations = 500,
  verbose = FALSE
)

# 5. Calculate alpha diversity
alpha_div <- colSums(ifelse(results > 0, 1, 0))
r_out <- r
values(r_out) <- alpha_div

# Visualize
par(mfrow=c(1,3))
plot(env_r, main = "Environment", col = terrain.colors(10))
plot(r_out, main = "Alpha Diversity", col = cm.colors(10))
plot(env_vals, alpha_div, xlab = "Environmental Value", ylab = "Alpha Diversity")

summary(lm(alpha_div ~ env_vals))
```

## 4. Spatial Feedbacks via Updater

In this example, we scale up to a spatial metacommunity (10x10 grid). We use the 
`updater` function to dynamically modify the environmental/spatial filter each epoch 
based on the spatial distribution of species abundances from the previous epoch. 

Specifically, cells where a species is highly abundant experience a locally lowered 
filter penalty (simulating habitat modification or niche construction).

```{r spatial-feedback, fig.width=8, fig.height=5}

# 1. Setup Spatial Landscape (10x10)
r <- rast(ncol=10, nrow=10, xmin=0, xmax=10, ymin=0, ymax=10)
C <- ncell(r)
S <- 50

# 2. Migration and Parameters
coords <- crds(r)
dists <- as.matrix(dist(coords))
M.migra <- 10^(-0.05 * dists)

Js <- rep(200, C)
Meta.pool <- runif(S)
Meta.pool <- Meta.pool / sum(Meta.pool)
d.spp <- rep(1, S)

# Initial uniform environmental filter (S x C matrix)
FF_init <- matrix(0.5, nrow = S, ncol = C)

# 3. Define the Spatial Updater Function
filter_updater <- function(current_comm, current_args) {
  
  # Ensure inputs are strictly matrices with preserved dimensions
  current_comm <- as.matrix(current_comm)
  current_FF <- as.matrix(current_args$FF)
  
  # Scale abundance safely as a matrix operation
  max_abund <- max(current_comm)
  if (max_abund > 0) {
    abund_scaled <- current_comm / max_abund
  } else {
    abund_scaled <- current_comm
  }
  # Ensure abund_scaled didn't lose its matrix status
  dim(abund_scaled) <- dim(current_comm)
  
  # Calculate filter
  new_FF <- current_FF + (0.5 * abund_scaled)
  
  # Clamp values using matrix-safe indexing
  new_FF[new_FF < 0.1] <- 0.1
  new_FF[new_FF > 1.0] <- 1.0
  
  # Final safety check on dimensions
  dim(new_FF) <- dim(current_FF)
  
  return(list(FF = new_FF))
}

# # 4. Run for exactly 1 Epoch
# results_one_epoch <- dynamicMetacomm(
#   nEpochs = 1,
#   Meta.pool = Meta.pool,
#   Js = Js,
#   M.migra = M.migra,
#   m.pool = 0.05,
#   d.spp = d.spp,
#   FF = FF_init,
#   updater = filter_updater,
#   nIterations = 500,
#   verbose = FALSE
# )
# 
# # 5. Use spatial_updater() to calculate the new filter state
# updated_args <- filter_updater(current_comm = results_one_epoch[[1]], current_args = list(FF = FF_init))
# 
# # 5. Plot Filter for two species
# par(mfrow = c(1, 2))
# 
# r_updated_filter <- r
# values(r_updated_filter) <- updated_args$FF[1, ]
# plot(r_updated_filter, main = "Filter (Epoch 1) for species 1", 
#      col = hcl.colors(10, "Reds", rev = TRUE), zlim = c(0.1, 1.0))
# 
# r_updated_filter <- r
# values(r_updated_filter) <- updated_args$FF[5, ]
# plot(r_updated_filter, main = "Filter (Epoch 1) for species 5", 
#      col = hcl.colors(10, "Reds", rev = TRUE), zlim = c(0.1, 1.0))


# 7. Run the Dynamic Spatial Simulation across 5 Epochs
results_spatial <- dynamicMetacomm(
  nEpochs = 10,
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = 0.01,
  d.spp = d.spp,
  FF = FF_init,          # Pass initial filter
  updater = filter_updater,
  nIterations = 500,
  verbose = FALSE
)

# 8. Compare Filter and Alpha Diversity at the Final Epoch
final_comm <- results_spatial[[length(results_spatial)]]
final_alpha <- colSums(ifelse(final_comm > 0, 1, 0))

r_div <- r
values(r_div) <- final_alpha

plot(r_div, main = "Alpha Diversity after Feedback Epochs", col = hcl.colors(10, "YlGnBu"))

# Plot dominant species across the landscape
r_dom <- list()
for(iii in 1:length(results_spatial)) {
  r_div <- r
  values(r_div) <- apply(results_spatial[[iii]], 2, which.max)
  r_dom[[iii]] <- r_div
}
r_dom <- rast(r_dom)
names(r_dom) <- paste0("Epoch_", 1:length(results_spatial))
plot(r_dom)

```

## 5. Mutualistic networks

This example demonstrates bipartite ecological assembly (e.g., plants and pollinators) 
using masterEqMetacommMutualistic(). Local recruitment probabilities in each guild are 
co-facilitated by species abundances from the interacting 
guild via a bipartite matrix alpha.animal.plant ($S_{animal} \times S_{plant}$).


```{r mutualistic, fig.width=8, fig.height=5}

# 1. Setup Spatial Landscape (10x10 Grid)
r <- rast(ncol = 10, nrow = 10, xmin = 0, xmax = 10, ymin = 0, ymax = 10)
C <- ncell(r)

coords <- crds(r)
dists <- as.matrix(dist(coords))
M.migra.plant <- 10^(-0.05 * dists)
diag(M.migra.plant) <- 0
M.migra.animal <- 10^(-0.02 * dists)

# 2. Define Guild Specifications (20 Animal Species, 30 Plant Species)
S_animal <- 50
S_plant  <- 50

Meta.pool.animal <- runif(S_animal)
Meta.pool.animal <- Meta.pool.animal / sum(Meta.pool.animal)

Meta.pool.plant <- runif(S_plant)
Meta.pool.plant <- Meta.pool.plant / sum(Meta.pool.plant)

Js.animal <- rep(400, C)
Js.plant  <- rep(400, C)

# 3. Environmental Filters
# Plant Filter: Sigmoidal suitability gradient centered at X = 5
x_coords <- coords[, 1] / 10 # Normalized 0 to 1 across X-axis
FF.plant <- matrix(0.001, nrow = S_plant, ncol = C)

for (i in 1:S_plant) {
  if (i <= 10) {
    # 10 Western Specialists
    FF.plant[i, ] <- ifelse(x_coords <= 5, 0.80, 0.001)
  } else if (i <= 20) {
    # 10 Eastern Specialists
    FF.plant[i, ] <- ifelse(x_coords > 5, 0.80, 0.001)
  } else {
    # 5 Generalists
    FF.plant[i, ] <- 0.05
  }
}

# Animal Filter: Uniformly harsh environment (0.1) across all patches
# Animals MUST have local mutualistic plant support to overcome this penalty
FF.animal <- matrix(1, nrow = S_animal, ncol = C)

# Ensure animal specialist suitabilities are balanced across species
FF.animal <- matrix(0.01, nrow = S_animal, ncol = C)
# If matching animal specialists to plant specialists:
for (j in 1:S_animal) {
  if (j <= 20) {
    FF.animal[j, ] <- ifelse(x_coords > 5, 0.80, 0.001) # Equal suitability for all animal specialists
  }
}

# 4. Create Strong Bipartite Facilitation Matrix
alpha.animal.plant <- matrix(rbinom(S_animal * S_plant, size = 1, prob = 0.1),
                             nrow = S_animal, ncol = S_plant)
# High interaction strengths amplify animal dependence on plant presence
alpha.animal.plant <- alpha.animal.plant * runif(S_animal * S_plant, 0.1, 0.5)

# 
# alpha.animal.plant <- matrix(0.01, nrow = S_animal, ncol = S_plant)
# diag(alpha.animal.plant) <- 100 # Animal i depends strictly on Plant i

plant_total_support <- colSums(alpha.animal.plant)
plant_total_support[plant_total_support == 0] <- 1
alpha.animal.plant <- sweep(alpha.animal.plant, 2, plant_total_support, FUN = "/") * 0.5

alpha.animal.plant <- matrix(0, nrow = S_animal, ncol = S_plant)

# 5. Simulate Dynamics with Strong Obligate Dependence
results_mut <- masterEqMutualistic(
  Meta.pool.animal   = Meta.pool.animal,
  Meta.pool.plant    = Meta.pool.plant,
  Js.animal          = Js.animal,
  Js.plant           = Js.plant,
  M.migra.animal     = M.migra.animal,
  M.migra.plant      = M.migra.plant,
  m.pool.animal      = 0.05,
  m.pool.plant       = 0.05,
  FF.animal          = FF.animal,
  FF.plant           = FF.plant,
  alpha.animal.plant = alpha.animal.plant,
  nIterations        = 500,
  verbose            = FALSE
)

# 6. Map Suitability Filter and Alpha Diversity for Both Guilds
mean_plant_suitability <- colMeans(FF.plant)
alpha_animal <- colSums(results_mut$Meta.animal > 0)
alpha_plant  <- colSums(results_mut$Meta.plant > 0)

alpha_animal <- rowSums(results_mut$Meta.animal)
alpha_plant  <- rowSums(results_mut$Meta.plant)

a <- results_mut$Meta.animal
rowSums(a)

b <- results_mut$Meta.plant
rowSums(b)

plot(Meta.pool.plant, log(rowSums(b)))
plot(rowMeans(FF.plant), log(rowSums(results_mut$Meta.plant)))


r_filter <- r_animal <- r_plant <- r
values(r_filter) <- mean_plant_suitability
values(r_plant)  <- alpha_plant
values(r_animal) <- alpha_animal

# 7. Visualize Filter and Spatial Guild Distributions
par(mfrow = c(2, 2))
plot(r_filter, main = "Mean Plant Filter (FF.plant)", col = hcl.colors(10, "Greens", rev = TRUE))
plot(r_plant, main = "Plant Guild Alpha Diversity", col = hcl.colors(10, "YlGn"))
plot(r_animal, main = "Animal Guild Alpha Diversity", col = hcl.colors(10, "Warm"))

plot(alpha_plant, alpha_animal, 
     xlab = "Plant Alpha Diversity", 
     ylab = "Animal Alpha Diversity",
     main = "4. Plant Richness Effect on Animals",
     pch = 19, col = adjustcolor("firebrick", alpha.f = 0.8), cex = 1.2)
summary(lm(alpha_animal ~ alpha_plant))

```

## 6. Landscape degradation

We simulate a disturbance event that reduces carrying capacity (`Js`) in the eastern half of the landscape (X >= 5) for a defined period of epochs. After the disturbance, we track the recovery of alpha diversity in both disturbed and undisturbed regions.

```{r dynamic-recolonization, fig.width=10, fig.height=8}

# 1. Setup Spatial Landscape (10x10)
r <- rast(ncol = 10, nrow = 10, xmin = 0, xmax = 10, ymin = 0, ymax = 10)
C <- ncell(r)
S <- 200

# Spatial coordinates & Distance-decay migration
coords <- crds(r)
dists  <- as.matrix(dist(coords))
M.migra <- 10^(-0.7 * dists)
diag(M.migra) <- 1.0

# Species & Environment Parameters
Js_baseline <- rep(50, C) 
Meta.pool   <- runif(S)
Meta.pool   <- Meta.pool / sum(Meta.pool)
d.spp       <- rep(1, S)


# Define East-half disturbance mask (X >= 5)
disturbance_mask <- coords[, 1] >= 5

# 2. Construct Js as a List Across Epochs (Disturbance Epochs 100-200)
nEpochs <- 500
start_disturbance <- 200
end_disturbance   <- 250

Js_list <- vector("list", nEpochs)

for (ep in 1:nEpochs) {
  if (ep >= start_disturbance && ep <= end_disturbance) {
    Js_epoch <- Js_baseline
    Js_epoch[disturbance_mask] <- 1  
    Js_list[[ep]] <- Js_epoch
  } else {
    Js_list[[ep]] <- Js_baseline
  }
}

# 3. Run Dynamic Metacommunity Simulation
results_recol <- dynamicMetacomm(
  nEpochs     = nEpochs,
  Meta.pool   = Meta.pool,
  Js          = Js_list,
  M.migra     = M.migra,
  m.pool      = 0.01,
  d.spp       = d.spp,
  nIterations = 1,
  verbose     = FALSE
)

# 4. Extract Alpha Across Epochs
epochs_count <- length(results_recol)

mean_alpha_disturbed   <- numeric(epochs_count)
mean_alpha_undisturbed <- numeric(epochs_count)

for (ep in 1:epochs_count) {
  comm <- results_recol[[ep]]
  
  # Alpha Diversity
  alpha <- colSums(comm > 0)
  mean_alpha_disturbed[ep]   <- mean(alpha[disturbance_mask])
  mean_alpha_undisturbed[ep] <- mean(alpha[!disturbance_mask])
  
}

# 6. Plots
plot(1:epochs_count, mean_alpha_disturbed, type = "l", col = "firebrick", lwd = 2,
     ylim = c(0, max(mean_alpha_undisturbed) * 1.25), 
     xlab = "Epoch", ylab = "Mean Alpha Diversity",
     main = "Alpha Diversity Trajectory")
lines(1:epochs_count, mean_alpha_undisturbed, col = "navy", lwd = 2, lty = 2)
rect(xleft = start_disturbance, ybottom = -5, xright = end_disturbance, ytop = 100, 
     col = adjustcolor("orange", alpha.f = 0.2), border = NA)
legend("topright", legend = c("Disturbed", "Undisturbed", "Disturbance time window" ),
       col = c("firebrick", "navy", adjustcolor("orange", alpha.f = 0.5)), 
       lty = c(1, 2, 1), lwd = c(2, 2, 6), bty = "n")

grid()




```

## 7. Change in spatial filter

```{r dynamic-spatial filter, fig.width=10, fig.height=8}

# 1. Setup Spatial Landscape (10x10)
r <- rast(ncol = 10, nrow = 10, xmin = 0, xmax = 10, ymin = 0, ymax = 10)
C <- ncell(r)
S <- 50
s_sens <- 20
s_tol  <- S - s_sens

# Spatial coordinates & Distance-decay migration
coords <- crds(r)
dists <- as.matrix(dist(coords))
M.migra <- 10^(-0.5 * dists)
diag(M.migra) <- 1.0

# Parameters
Js_baseline <- rep(50, C)
Meta.pool   <- runif(S)
Meta.pool   <- Meta.pool / sum(Meta.pool)
d.spp       <- rep(1, S)

# Define East-half disturbance mask (X >= 5) and species groups
disturbance_mask <- coords[, 1] >= 5
sensitive_spp    <- 1:s_sens
tolerant_spp     <- s_tol:S

# 2. Construct FF as a List of Matrices Across Epochs
nEpochs <- 500
start_disturbance <- 100
end_disturbance   <- 200

FF_baseline <- matrix(0.5, nrow = S, ncol = C)
FF_list     <- vector("list", nEpochs)

for (ep in 1:nEpochs) {
  if (ep >= start_disturbance && ep <= end_disturbance) {
    FF_epoch <- FF_baseline
    # Reduce environmental suitability ONLY for sensitive species in East zone
    FF_epoch[sensitive_spp, disturbance_mask] <- 0.05
    FF_list[[ep]] <- FF_epoch
  } else {
    FF_list[[ep]] <- FF_baseline
  }
}

# 3. Run Dynamic Metacommunity Simulation (Passing FF_list directly)
results_ff <- dynamicMetacomm(
  nEpochs     = nEpochs,
  Meta.pool   = Meta.pool,
  Js          = Js_baseline,   # Static carrying capacity
  M.migra     = M.migra,
  m.pool      = 0.001,
  d.spp       = d.spp,
  FF          = FF_list,       # Time-varying filter matrix list
  nIterations = 1,
  verbose     = FALSE
)

# 4. Extract Species Group Abundances in the Disturbed East Zone
epochs_count <- length(results_ff)
mean_abund_sensitive <- numeric(epochs_count)
mean_abund_tolerant  <- numeric(epochs_count)

for (ep in 1:epochs_count) {
  comm <- results_ff[[ep]]
  
  # Average abundance per cell in East zone
  mean_abund_sensitive[ep] <- mean(colSums(comm[sensitive_spp, disturbance_mask, drop = FALSE]))
  mean_abund_tolerant[ep]  <- mean(colSums(comm[tolerant_spp, disturbance_mask, drop = FALSE]))
}

# Plot
max_y <- max(c(mean_abund_sensitive, mean_abund_tolerant)) * 1.2
plot(1:epochs_count, mean_abund_sensitive, type = "l", col = "firebrick", lwd = 2,
     ylim = c(0, max_y), xlab = "Epoch", ylab = "Mean Abundance per Cell (East Zone)",
     main = "Response to Dynamic FF Filter")
lines(1:epochs_count, mean_abund_tolerant, col = "forestgreen", lwd = 2, lty = 2)

rect(xleft = start_disturbance, ybottom = -5, xright = end_disturbance, ytop = max_y * 1.1, 
     col = adjustcolor("orange", alpha.f = 0.2), border = NA)

legend("topright", legend = c("Sensitive species", "Tolerant species", "Filter Drop Window"),
       col = c("firebrick", "forestgreen", adjustcolor("orange", alpha.f = 0.5)), 
       lty = c(1, 2, 1), lwd = c(2, 2, 6), bty = "n")
grid()

```










