---
title: "examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Examples of dynamet application

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r packages }
library(terra)
library(dynamet)
library(dplyr)
library(ggplot2)
```


## 1. Single Community Assembly

To understand the core lottery dynamics, we start with a single local community ($C=1$)
receiving individuals from a regional pool of 5 species.

```{r single-community}
# 1. Setup: 10 species in the regional pool
S <- 5
Meta.pool <- c(0.1, 0.2, 0.3, 0.3, 0.1) # Relative abundances

# 2. Parameters for a single community
# Js: Carrying capacity of 500 individuals
# M.migra: 1x1 matrix (closed system relative to other communities)
Js <- 500
M.migra <- matrix(1, nrow = 1, ncol = 1)
m.pool <- 0.05  # 5% immigration rate from the meta-pool
d.spp <- rep(1, S) # Uniform dispersal ability across species
nEpochs = 20 # Set the number of epochs

# 3. Initialize community with zero individuals
init_comm <- matrix(0, nrow = S, ncol = 1)

# 4. Run for 10 epochs
results <- dynamicMetacomm(
  nEpochs = nEpochs,
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = m.pool,
  d.spp = d.spp,
  lottery = TRUE,
  nIterations = 200,
  prop.dead.by.it = 0.1,
  Ea = 0.5,
  init.comm = init_comm,
  verbose = FALSE
)

# 5. Visualize the transition from zero to steady-state
sad_data <- do.call(rbind, lapply(1:nEpochs, function(i) {
  data.frame(
    Epoch = i,
    Species = factor(1:S),
    Abundance = as.vector(results[[i]])
  )
}))

# Plot: Stacked bar chart showing relative composition over time
ggplot(sad_data, aes(x = factor(Epoch), y = Abundance, fill = Species)) +
  geom_bar(stat = "identity", position = "fill") + # "fill" shows relative proportion
  labs(
    title = "Species Abundance Distribution (SAD) Over Time",
    x = "Epoch",
    y = "Relative Abundance"
  ) +
  theme_minimal()
```

## 2. Spatial Metacommunity Simulation

In this example, we model 100 communities arranged on a 10x10 grid, where carrying 
capacity (`Js`) varies spatially and migration is constrained by distance.

```{r spatial-sim, fig.width=8, fig.height=4}

# 1. Define the 10x10 Landscape
r <- rast(ncol=10, nrow=10, xmin=0, xmax=10, ymin=0, ymax=10)

# 2. Create spatially structured Carrying Capacity (Js)
values(r) <- 50 + 50 * sin(seq(0, pi, length.out=100))
Js <- as.vector(floor(values(r)))

# 3. Create Migration Matrix (M.migra)
coords <- crds(r)
dists <- as.matrix(dist(coords))
M.migra <- 10^(-0.1 * dists)

# 4. Parameters
S <- 50
Meta.pool <- runif(S)
d.spp <- runif(S)

# 5. Run the Simulation
# Using masterEqMetacomm for a single epoch/iteration
community_abundance <- masterEqMetacomm(
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = 0.05,
  d.spp = d.spp,
  nIterations = 500,
  verbose = FALSE
)

# 6. Map results back to raster for visualization
alpha_div <- colSums(community_abundance > 0)
r_out <- r
values(r_out) <- alpha_div

# Visualize
par(mfrow=c(1,2))
plot(r, main = "Carrying Capacity (Js)", col = terrain.colors(10))
plot(r_out, main = "Alpha Diversity", col = cm.colors(10))
```

## 3. Spatial Metacommunity Simulation with a spatial filter

In this example, we model 100 communitieswith 20 species arranged on a 10x10 grid, where carrying 
capacity (`Js`) varies spatially and migration is constrained by distance.

```{r spatial-sim-filter, fig.width=8, fig.height=4}

# 1. Define the 10x10 Landscape Environment (e.g., elevation or pH)
env_r <- r <- rast(ncol=10, nrow=10, xmin=0, xmax=10, ymin=0, ymax=10)
values(env_r) <- seq(0, 1, length.out=100) # Simple gradient across the landscape
values(r) <- 50 + 50 * sin(seq(0, pi, length.out=100))
Js <- as.vector(floor(values(r)))

# 2. Define Species Niche Optima (S = 5)
S <- 50
niche_optima <- seq(0.75, 1, length.out = S)

# 3. Create the Environmental Filtering Matrix (FF)
# FF must be S x C (5 rows x 100 columns)
# We calculate a Gaussian penalty for each species in each cell
FF <- matrix(NA, nrow = S, ncol = 100)
env_vals <- values(env_r)

for(i in 1:S) {
  # Gaussian filter: 1 = perfect match, approaching 0 = high penalty
  FF[i, ] <- exp(-((env_vals - niche_optima[i])^2) / (2 * 0.2^2))
}

# 4. Run the simulation with Filtering
# Note: FF is now passed as a matrix of dimensions S x C
results <- masterEqMetacomm(
  Meta.pool = rep(1/S, S),
  Js = Js,
  M.migra = M.migra, # Assuming M.migra is defined as in the previous step
  m.pool = 0.05,
  d.spp = rep(1, S),
  FF = FF,           # Spatial Filter Applied
  nIterations = 500,
  verbose = FALSE
)

# 5. Visualize a species' distribution filtered by the environment
alpha_div <- colSums(ifelse(results > 0, 1, 0))
r_out <- r
values(r_out) <- alpha_div

# Visualize
par(mfrow=c(1,2))
plot(env_r, main = "Environment", col = terrain.colors(10))
plot(r_out, main = "Alpha Diversity", col = cm.colors(10))
```

## 4. Advanced: Spatial Eco-Evolutionary Feedbacks via Updater

In this example, we scale up to a spatial metacommunity (10x10 grid). We use the 
`updater` function to dynamically modify the environmental/spatial filter each epoch 
based on the spatial distribution of species abundances from the previous epoch. 

Specifically, cells where a species is highly abundant experience a locally lowered 
filter penalty (simulating habitat modification or niche construction).

```{r spatial-feedback, fig.width=8, fig.height=5}

# 1. Setup Spatial Landscape (10x10)
r <- rast(ncol=10, nrow=10, xmin=0, xmax=10, ymin=0, ymax=10)
C <- ncell(r)
S <- 50

# 2. Migration and Parameters
coords <- crds(r)
dists <- as.matrix(dist(coords))
M.migra <- 10^(-0.1 * dists)

Js <- rep(200, C)
Meta.pool <- runif(S)
Meta.pool <- Meta.pool / sum(Meta.pool)
d.spp <- rep(1, S)

# Initial uniform environmental filter (S x C matrix)
FF_init <- matrix(0.5, nrow = S, ncol = C)

# 3. Define the Spatial Updater Function
filter_updater <- function(current_comm, current_args) {
  
  # Ensure inputs are strictly matrices with preserved dimensions
  current_comm <- as.matrix(current_comm)
  current_FF <- as.matrix(current_args$FF)
  
  # Scale abundance safely as a matrix operation
  max_abund <- max(current_comm)
  if (max_abund > 0) {
    abund_scaled <- current_comm / max_abund
  } else {
    abund_scaled <- current_comm
  }
  # Ensure abund_scaled didn't lose its matrix status
  dim(abund_scaled) <- dim(current_comm)
  
  # Calculate filter
  new_FF <- current_FF - (0.5 * abund_scaled)
  
  # Clamp values using matrix-safe indexing
  new_FF[new_FF < 0.1] <- 0.1
  new_FF[new_FF > 1.0] <- 1.0
  
  # Final safety check on dimensions
  dim(new_FF) <- dim(current_FF)
  
  return(list(FF = new_FF))
}

# 4. Run for exactly 1 Epoch
results_one_epoch <- dynamicMetacomm(
  nEpochs = 1,
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = 0.01,
  d.spp = d.spp,
  FF = FF_init,
  updater = filter_updater,
  nIterations = 500,
  verbose = FALSE
)

# 5. Use spatial_updater() to calculate the new filter state
updated_args <- filter_updater(current_comm = results_one_epoch[[1]], current_args = list(FF = FF_init))

# 5. Plot Filter for two species
par(mfrow = c(1, 2))

r_updated_filter <- r
values(r_updated_filter) <- updated_args$FF[1, ]
plot(r_updated_filter, main = "Filter (Epoch 1) for species 1", 
     col = hcl.colors(10, "Reds", rev = TRUE), zlim = c(0.1, 1.0))

r_updated_filter <- r
values(r_updated_filter) <- updated_args$FF[5, ]
plot(r_updated_filter, main = "Filter (Epoch 1) for species 5", 
     col = hcl.colors(10, "Reds", rev = TRUE), zlim = c(0.1, 1.0))


# 7. Run the Dynamic Spatial Simulation across 5 Epochs
results_spatial <- dynamicMetacomm(
  nEpochs = 10,
  Meta.pool = Meta.pool,
  Js = Js,
  M.migra = M.migra,
  m.pool = 0.01,
  d.spp = d.spp,
  FF = FF_init,          # Pass initial filter
  updater = filter_updater,
  nIterations = 500,
  verbose = FALSE
)

# 8. Compare Filter and Alpha Diversity at the Final Epoch
final_comm <- results_spatial[[length(results_spatial)]]
final_alpha <- colSums(ifelse(final_comm > 0, 1, 0))

r_div <- r
values(r_div) <- final_alpha

plot(r_div, main = "Alpha Diversity after Feedback Epochs", col = hcl.colors(10, "YlGnBu"))

# Plot dominant species across the landscape
r_dom <- list()
for(iii in 1:length(results_spatial)) {
  r_div <- r
  values(r_div) <- apply(results_spatial[[iii]], 2, which.max)
  r_dom[[iii]] <- r_div
}
r_dom <- rast(r_dom)
names(r_dom) <- paste0("Epoch_", 1:length(results_spatial))
plot(r_dom)

```

## 5. Advanced: Spatial Eco-Evolutionary Feedbacks via Updater

This example demonstrates bipartite ecological assembly (e.g., plants and pollinators) 
using masterEqMetacommMutualistic(). Local recruitment probabilities in each guild are 
co-facilitated by species abundances from the interacting 
guild via a bipartite matrix alpha.animal.plant ($S_{animal} \times S_{plant}$).


```{r mutualistic, fig.width=8, fig.height=5}

# 1. Setup Spatial Landscape (10x10 Grid)
r <- rast(ncol = 10, nrow = 10, xmin = 0, xmax = 10, ymin = 0, ymax = 10)
C <- ncell(r)

coords <- crds(r)
dists <- as.matrix(dist(coords))
M.migra <- 10^(-0.1 * dists)

# 2. Define Guild Specifications (15 Animal Species, 25 Plant Species)
S_animal <- 30
S_plant  <- 25

Meta.pool.animal <- runif(S_animal)
Meta.pool.plant <- runif(S_plant)

Js.animal <- rep(50, C)
Js.plant  <- rep(100, C)

# 3. Create Bipartite Mutualistic Matrix (S_animal x S_plant)
alpha.animal.plant <- matrix(
  rbinom(S_animal * S_plant, size = 1, prob = 0.25),
                             nrow = S_animal, ncol = S_plant)
alpha.animal.plant <- alpha.animal.plant * runif(S_animal * S_plant, 0.1, 0.5)

# 4. Simulate Mutualistic Metacommunity Dynamics
results_mut <- masterEqMutualistic(
  Meta.pool.animal = Meta.pool.animal,
  Meta.pool.plant  = Meta.pool.plant,
  Js.animal        = Js.animal,
  Js.plant         = Js.plant,
  M.migra.animal   = M.migra,
  M.migra.plant    = M.migra,
  m.pool.animal    = 0.05,
  m.pool.plant     = 0.05,
  alpha.animal.plant = alpha.animal.plant,
  nIterations      = 300,
  verbose          = FALSE
)

# 5. Map Alpha Diversity for Both Guilds
alpha_animal <- colSums(results_mut$Meta.animal > 0)
alpha_plant  <- colSums(results_mut$Meta.plant > 0)

r_animal <- r
values(r_animal) <- alpha_animal

r_plant <- r
values(r_plant) <- alpha_plant

# Visualize Spatial Guild Distributions
par(mfrow = c(1, 2))
plot(r_animal, main = "Animal Alpha Diversity", col = hcl.colors(10, "Warm"))
plot(r_plant, main = "Plant Alpha Diversity", col = hcl.colors(10, "YlGn"))

plot(r_animal, r_plant)


```

